---
title: "Homework Assignment 1"
author: "Michael Kalmus"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
subtitle: 'https://github.com/mkalmus/TO628-Fall-2020-Work'
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: 
      collapsed: false
---

## HW-1 Instructions

> Our class emphasizes clear communication of data analysis results to non-technical audiences. I expect your HTML output documents to be readable and well formatted. I expect you to put ample comments in your R code to make the code understandable. Along with accuracy of results, I am looking for quality of presentation as well. This homework is due by **11:59PM on Nov 01st**.  To complete this assignment, follow these steps:

1. Create a new RStudio Project for this HW. Sync the project directory with a GitHub Repository (see instructions on Canvas on how to do that). 

2. Download the `HW1.Rmd` file from Canvas. Save the file in the RStudio Project Directory. 

3. Open `HW1.Rmd` in RStudio. Replace the "Your Name Here" text in the `author:` field with your name.

4. Supply your solutions to the homework by editing `HW1.Rmd`.

5. Run your code in the Console and Knit HTML frequently to check for errors.

6. You may find it easier to solve a problem by interacting only with the Console at first. 

7. When you have completed the homework please check that your knits correctly when you click `Knit HTML`.

8. "Push" all your local changes to the GitHub Repo for the project.

9. Submit your RMD file, the HTML output and a link to your GitHub Repo on Canvas.


## Data frame basics

We will continue working with the nycflights dataset we looked at in class. **Please be sure to keep the data file in the same directory as the RMD file - your project directory.**

First - make sure that you are reading the data correctly and doing all the data cleaning steps that we did in class. Feel free to go beyond our class work in cleaning the data if you feel the need. 

I now would like you to answer the following - all in their own separate R code blocks.

### Data Exploration

Let's first do some simple exploration of this data. 

```{r}
library(dplyr)
library(ggplot2)
```

```{r}
# Load data
flights <- read.csv("nycflights.csv")
```

```{r}
# Remove extra indicator column named X
flights$X <- NULL

# Factorize columns that need to be a factor
flights$origin <- as.factor(flights$origin)
flights$tailnum <- as.factor(flights$tailnum)
flights$month <- as.factor(flights$month)
flights$dest <- as.factor(flights$dest)
flights$carrier <- as.factor(flights$carrier)
flights$flight <- as.factor(flights$flight)

# Re-check data
str(flights)
```

- How many airlines are there? (Hint: `levels` and `length` can be useful here)

```{r}
length(levels(flights$carrier))
```
There are `r length(levels(flights$carrier))` airlines

- How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)

```{r}
flights %>% filter(carrier == "OO") %>% nrow()
```
There are `r flights %>% filter(carrier == "OO") %>% nrow()` flights with airline code 'OO'

- How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)

```{r}
min(flights$air_time, na.rm = TRUE)
```

The shortest flight out of any NYC airport is `r min(flights$air_time, na.rm = TRUE)` minutes.

- How many flights where there by United Airlines (code: UA) on Jan 12th 2013?

```{r}
flights %>% filter(carrier == "UA" & year == "2013" & day == "12" & month == "1") %>% nrow()
```
There were `r flights %>% filter(carrier == "UA" & year == "2013" & day == "12" & month == "1") %>% nrow()` flights by United Airlines on Jan 12th, 2013.

### Arrival Delay

Lets focus on Arrival Delay.

- What was the average arrival delay for all airports and all airlines combined in Jan 2013? 

```{r}
flights %>%
  filter(month == "1" & year == "2013") %>% 
  select(arr_delay) %>%
  summarise(arr = mean(arr_delay, na.rm = TRUE))
```
The average arrival delay for all airports and all airlines combined in January 2013 was 6.13 minutes.

- Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?

```{r}
flights %>%
  filter(month == "1" & year == "2013") %>% 
  select(arr_delay) %>%
  summarise(arr = median(arr_delay, na.rm = TRUE))
```
The median arrival delay for all airports and all airlines combined in Jan 2013 was -3 minutes, meaning planes left early.

Based on your answers to the two questions above, what can you say about the distribution of arrival delays? Provide your answer in a text paragraph form.

```{r}
flights %>%
  filter(month == "1" & year == "2013") %>% 
  select(arr_delay) %>%
  ggplot() + geom_histogram(aes(x = arr_delay), color = "red", bins = 50, na.rm = TRUE) + xlim(-50, 250)
```

Since the mean is greater than the median, the data for arrival delays is skewed to the right.

### Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned. For this question you will have to make sure that airline column is coded as a factor.

- Calculate average arrival delays by airline (Hint: look up the command `tapply`)

```{r}
# Using dplyr
carrier_delays_dplyr <- flights %>% 
  group_by(carrier) %>% 
  summarize(mean_delay = mean(arr_delay, na.rm = TRUE, .groups = "drop")) %>% 
  arrange(desc(mean_delay))

# Using base R
carrier_delays_base <- tapply(flights$arr_delay, flights$carrier, mean, na.rm = TRUE)

carrier_delays_dplyr
carrier_delays_base
```

- Draw a Bar Plot of Average Arrival Delays for all the Airlines (Hint: command for making a Bar Plot is simply `barplot`)

```{r}
# Base R
barplot(carrier_delays_base)

# dplyr/ggplot
carrier_delays_dplyr %>% ggplot() + geom_col(aes(x = carrier, y = mean_delay), color = "black") + theme_classic()
```

- Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? Provide answer to this question in a text paragraph form using **inline R code**.

```{r}
#using dplyr
carrier_delays_dplyr %>% filter(mean_delay == max(mean_delay)) %>% select(carrier) 
carrier_delays_dplyr %>% filter(mean_delay == min(mean_delay)) %>% select(carrier)
carrier_delays_dplyr %>% filter(mean_delay < 0) %>% select(carrier)
carrier_delays_dplyr$carrier[carrier_delays_dplyr$mean_delay == max(carrier_delays_dplyr$mean_delay)]

#using base R
names(carrier_delays_base[carrier_delays_base == max(carrier_delays_base)])
names(carrier_delays_base[carrier_delays_base == min(carrier_delays_base)])
names(carrier_delays_base[carrier_delays_base < 0])
```
`r names(carrier_delays_base[carrier_delays_base == max(carrier_delays_base)])` has the highest average arrival delay.
`r names(carrier_delays_base[carrier_delays_base == min(carrier_delays_base)])` has the lowest average arrival delay.
`r names(carrier_delays_base[carrier_delays_base < 0])` have negative average arrival delays, so yes airlines have negative average delays.

### Air Gain

Create a new column named airgain such that airgain = (departure delay - arrival delay) : this is the amount of delay a flight made up while in air.

```{r}
flights <- flights %>% mutate(airgain = dep_delay - arr_delay)
```


a) Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.

```{r}
min(flights$airgain, na.rm = TRUE)
max(flights$airgain, na.rm = TRUE)
mean(flights$airgain, na.rm = TRUE)
median(flights$airgain, na.rm = TRUE)
summary(flights$airgain)
flights %>% ggplot() + geom_boxplot(aes(x = carrier, y = airgain))
```


b) Answer the questions:

  - do airlines actually gain any time when in air on average? 
  
```{r}
mean(flights$airgain, na.rm = TRUE)
```

Yes, airlines actually gain ```r mean(flights$airgain, na.rm = TRUE)``` time when in air on average.

  - Calculate average airgain for different airlines - which airlines do a better job, which do a worse job? 
  
```{r}
flights %>% 
  group_by(carrier) %>% 
  summarize(mean_gain = mean(airgain, na.rm = TRUE)) %>% 
  arrange(desc(mean_gain)) 

```

From the dataframe, we can see that carrier AS does the best job, gaining 15.76 mins of airtime on average, while F9 does the worst, losing 1.72 mins of airtime on average.

Assuming a "better job" is gaining  more than 10 minutes of airtimes, carriers AS, HA, and VX do a better job. Conversely, if a worse job is losing airtime, carriers MQ, FL, and F9 do a worse job. 

### Merging Data Frames

> This section and the next is new compared to the class exercise. As you had an opportunity to work together in your breakout rooms for previous questions, this and the next section will carry a higher weight in grading for this HW.

You can get detailed information about the physical planes in our dataset in this file: `planes.csv`. Download and save this file in your project directory.

a) Read the `planes.csv` file using `read.csv` command. Do any data cleaning necessary.

```{r}
planes <- read.csv("planes.csv")
```

```{r}
planes$X <- NULL
planes$tailnum <- factor(planes$tailnum)
planes$type <- factor(planes$type)
planes$manufacturer <- factor(planes$manufacturer)
planes$model <- factor(planes$model)
planes$engine <- factor(planes$engine)
planes$year <- factor(planes$year)

str(planes)
```

b) Merge the flights data and the planes data using the `merge` command. You should do the merge on the common column named `tailnum`. *getting this right may need some trial and error and getting some help*.

```{r}
merge_df <- merge(flights, planes, sort = FALSE, by = "tailnum")
merge_df$X <- NULL
```

c) Now that you have a merged dataset, think of what interesting questions that you can ask that can be answered using the merged dataset. You are asked to pose five interesting questions and answer them. (For example: who are the top 10 manufacturers of planes that fly out of NYC airports?) **Be creative. Be bold. Ask questions that you would want to know answers to even if you were not doing this for a HW. **

Q1: How does average arrival delay vary by engine type?
```{r}
merge_df$engine <- factor(merge_df$engine)

q1df <- merge_df %>% 
  group_by(engine) %>% 
  summarize(mean_arr_delay = mean(arr_delay, na.rm = TRUE), .groups = "keep") %>% 
  arrange(desc(mean_arr_delay))
q1df

q1df %>% ggplot() + geom_col(aes(x = engine, y = mean_arr_delay), na.rm = TRUE)
```
Based on the tabular output and bar chart, it is clear that 4 Cycle engines have the highest average arrival delay and Turbo-jet engines have the lowest average arrival delay. Furthermore, the average arrival delay for 4 Cycle engines is more than 3 times greater than that of Turbo-jets. 


Q2: How do the amount of flights vary by month for the top 2 manufacturers?
```{r}
#find top 2 manufacturers
merge_df %>% group_by(manufacturer) %>% summarize(count = n(), .groups = "drop") %>% arrange(desc(count)) %>% slice(1:2)

#table output
merge_df %>% 
  filter(manufacturer == c("BOEING", "EMBRAER")) %>% 
  group_by(manufacturer, year.x, month) %>% 
  summarize(flights = n(), .groups = "drop")

#visualization
merge_df %>% 
  filter(manufacturer == c("BOEING", "EMBRAER")) %>% 
  group_by(manufacturer, year.x, month) %>% 
  summarize(flights = n(), .groups = "drop") %>% 
  ggplot() + 
  geom_col(aes(x = month, y = flights)) + facet_grid(~ manufacturer)
```
First, we find the top 2 manufacturers based on number of flights are Boeing and Embraer. The barchart shows how each varied per month in 2013, and we can see that variation exists in the number of flights per month for both manufacturers. Both manufacturers had their minimum flights in February, and the average appears to be the highest between May and September. Practically, this makes sense as many people fly over the summer. When looking at the number of flights between manufacturers, Boeing has on average more flights than Embraer in each month.


Q3: How does average arrival and departure delays vary by number of seats and does a correlation exist between seats and delays?
```{r}
merge_df %>% 
  group_by(seats) %>% 
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE), avg_dep_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>% 
  arrange(desc(avg_arr_delay))

merge_df %>% 
  group_by(seats) %>% 
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE), avg_dep_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>% 
  ggplot() + geom_col(aes(x = seats, y = avg_arr_delay))

merge_df %>% 
  group_by(seats) %>% 
  summarize(avg_arr_delay = mean(arr_delay, na.rm = TRUE), avg_dep_delay = mean(dep_delay, na.rm = TRUE), .groups = "drop") %>% 
  ggplot() + geom_col(aes(x = seats, y = avg_dep_delay))

cor(merge_df$seats, merge_df$arr_delay, use = "complete.obs")
cor(merge_df$seats, merge_df$dep_delay, use = "complete.obs")
```
Based on the output table and charts, no significant relationship seems to exist between the number of seats and average departure delay or average arrival delay. This was validated by finding the correlation coefficients between seats and each delay, which showed that both delay types had near-zero correlations with the number of seats. 


Q4: We know arrival delay and departure delay are not correlated with the number of seats: Are arrival delay and departure delay correlated with each other overall and does this change by plane type?
```{r}
#overall correlation
cor(merge_df$arr_delay, merge_df$dep_delay, use = "complete.obs")

#correlation by airport
merge_df %>%
  group_by(type) %>% 
  summarize(correlation = cor(arr_delay, dep_delay, use = "complete.obs"))
```

Q5: What manufacturer had the most air time and from which origin airport did the most air time come from? 
```{r}
#find manufacturer with the most air time
merge_df %>% 
  group_by(manufacturer) %>% 
  summarize(total_air_time = sum(air_time, na.rm = TRUE), .groups = "drop") %>% 
  arrange(desc(total_air_time)) %>% 
  slice(1)

merge_df %>% 
  filter(manufacturer == "BOEING") %>% 
  group_by(manufacturer, origin) %>% 
  summarize(total_air_time = sum(air_time, na.rm = TRUE), .groups = "drop") %>% 
  slice(1)
```
First, we see that Boeing had the most airtime in 2013 with a total air time of 17767056 minutes. When looking at Boeing by origin airport, we can see the origin airport with the most air time for Boeing was EWR with a total air time of 8444440 minutes.

### Making Your HTML Look Nice

We want our report to be good looking, professional documents. To that end, I am asking you to do the following:

- Have a floating table of contents

- Include code folding in your output. You can find more about code folding here: https://bookdown.org/yihui/rmarkdown/html-document.html#code-folding


That's it. Once you are done, make sure everything works and knits well and then you can push your changes to the GitHub repo and uplaod the RMD flile and the html output to Canvas.

**Have Fun!**

Sanjeev
